Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-08 introduction.tex
Mahsa's suggestions.

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
weak isolation level used in the experiments. The same study has found
that 13\% of all invariants among 67 open source Ruby-on-Rails
applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
been echoed by database researchers who lament the lack of a
better understanding of this problem; see e.g., \url{
  http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

>>>>

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under \emph{Read
Committed} weak isolation level - default level for 8 of the 18
databases studied in~\cite{bailishotos}, and found that 25 out of
every 1000 rows in the database violate at least one integrity
constraint. Bailis \emph{et al.}~\cite{bailisferal} rely on Rails'
\emph{uniqueness validation} to maintain uniqueness of records while
serving Linkbench's~\cite{linkbench} insertion workload (6400 records
distributed over 1000 keys; 64 concurrent clients), and report
discovering more than 10 duplicate records.  Rails relies on database
transactions to validate uniqueness during insertions, which is
sensible if transactions are serializable, but incorrect under the
weak isolation level used in the experiments. The same study has found
that 13\% of all invariants among 67 open source Ruby-on-Rails
applications are at risk of being violated due to weak isolation.
Indeed, incidents of safety violations due to executing applications
in a weakly-isolated environment have been reported on web services in
production~\cite{starbucksbug, scimedbug}, including in
safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
been echoed by database researchers who lament the lack of a better
understanding of this problem; see e.g., \url{
  http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

### 2017-11-08 motivation.tex
Mahsa's suggestions.

This example does not exhibit any of the anomalies that
\emph{characterize} RC isolation~\cite{berenson}\footnote{Berenson et
al characterize isolation levels in terms of the
anomalies they \emph{admit}. For example, RC is characterized by
\emph{lost writes} because it admits the anomaly.}. For instance,
there are no \emph{lost writes} since both concurrent transactions' writes
are present in the final state of the database.  Program analyses that
aim to determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
can complicate high-level reasoning.  A possible alternative would be
to utilize concurrent program verification methods where the
implementation details. of weak isolation are interleaved within the
program, yielding a (more-or-less) conventional concurrent program;
but, considering the size and complexity of real-world transaction
systems, this strategy is unlikely to scale.

@@

This observation leads us away from thinking of a collection of
database transactions as a simple variant of a concurrent imperative
program.  Instead, we see value in viewing them as essentially
functional computations that manage database state abstractly,
mirroring the structure of our DSL.  By doing so, we can formulate the
semantics of database operations as state transformers that explicitly
relate an operation's pre- and post-states, defining the semantics of
the corresponding transformer algorithmically, just like classical
predicate transformer semantics (e.g., weakest pre-condition or
strongest post-condition).  In our case, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
records, and a SQL statement as a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument (i.e., the loop
body) to the state transformer of the combined expression (i.e., the
loop).  We illustrate this intuition on a simple example.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{Bind ($\bind$) has higher precedence than union
($\cup$). Angle braces ($\langle \ldots \rangle$) are used to denote
records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda \C{s}. 
      \itel{{\sf table}(\C{s}) = \C{Stock} \conj 
        \C{s}.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.15in}}
           {\{ \langle \C{s\_i\_id}=\C{s.s\_i\_id};\, 
                       \C{s\_d\_id}=\C{s.s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.15in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s\_qty}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to $\stl$ is not already present in $\stl$.}

>>>>

This example does not exhibit any of the anomalies that
\emph{characterize} RC isolation~\cite{berenson}\footnote{Berenson
\emph{et al.} characterize isolation levels in terms of the anomalies
they \emph{admit}. For example, RC is characterized by \emph{lost
writes} because it admits the anomaly.}. For instance, there are no
\emph{lost writes} since both concurrent transactions' writes are
present in the final state of the database.  Program analyses that aim
to determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
can complicate high-level reasoning.  A possible alternative would be
to utilize concurrent program verification methods where the
implementation details of weak isolation are interleaved within the
program, yielding a (more-or-less) conventional concurrent program.
But, considering the size and complexity of real-world transaction
systems, this strategy is unlikely to scale.

@@

This observation leads us away from thinking of a collection of
database transactions as a simple variant of a concurrent imperative
program.  Instead, we see value in viewing them as essentially
functional computations that manage database state abstractly,
mirroring the structure of our DSL.  By doing so, we can formulate the
semantics of database operations as state transformers that explicitly
relate an operation's pre- and post-states, defining the semantics of
the corresponding transformer algorithmically, just like classical
predicate transformer semantics (e.g., weakest pre-condition or
strongest post-condition).  In our case, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
records, and a SQL statement as a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument, i.e., the loop
body, to the state transformer of the combined expression, i.e., the
loop.  We illustrate this intuition on a simple example.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1}, \C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{Bind ($\bind$) has higher precedence than union
($\cup$). Angle braces ($\langle \ldots \rangle$) are used to denote
records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda \C{s}. 
      \itel{{\sf table}(\C{s}) = \C{Stock} \conj 
        \C{s}.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.15in}}
           {\{ \langle \C{s\_i\_id}=\C{s.s\_i\_id};\, 
                       \C{s\_d\_id}=\C{s.s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.15in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s\_qty}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to $\stl$ is not already present in $\stl$.}

### 2017-11-08 opsem.tex
Minor edits.
Mahsa's suggestions.

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\langle \bar{f}=\bar{e} \rangle$) constitute the
syntactic class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$),
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}).  Our language ensures
that all effectful actions are encapsulated within database commands,
and that all shared state among processes are only manipulated via
transactions and its supported operations.  In particular, we do not
consider programs in which objects resident on e.g., the OCaml heap
are concurrently manipulated by OCaml expressions as well as database
actions.

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), Integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\langle \bar{f}=\bar{e} \rangle$) constitute the
syntactic class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$),
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}). Our language ensures
that all effectful actions are encapsulated within database commands,
and that all shared state among processes are only manipulated via
transactions and its supported operations.  In particular, we do not
consider programs in which objects resident on e.g., the OCaml heap
are concurrently manipulated by OCaml expressions as well as database
actions.
